---
title: "gmse: Generalised management strategy evaluation"
author: "A. Bradley Duthie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: gmse.bib
---

```{r, echo = FALSE}
library(GMSE);
```

# Introduction

- Broad paragraph about managing populations and conservation given uncertainty and conflict among stakeholders.

- Paragraph introducing adaptive management, then management strategy evaluation framework [@Bunnefeld2011]; introduce this *generalised* management strategy evaluation framework and R package as filling an important gap in tools available for modelling mangement.

- More details about what types of problems *GMSE* was developed to solve and (very generally) how it can solve them through individual-based modelling and a game-theoretic approach that uses genetic algorithms to simulate manager and stakeholder decision making.

# GMSE model structure

Introduce the general mode of generalised management strategy evaluation (e.g., below), and the idea of simulating the package simulating the whole process of management.

```{r, echo = FALSE, fig.height = 4, fig.width = 4}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=55, labels="Genetic", cex=1.0);
text(x=50, y=48, labels="algorithm", cex=1.0);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Off-take", srt=-90, cex = 1.0);
```

**Figure 1:** *Description of the generalised management strategy evaluation framework*

Next there will be some general explanation of what it is that each sub-model does. The subfunction called in R will be mentioned, but with emphasis that the software user need not do this themselves. I will emphasise that each model can be built upon as desired to include more (or less) detail.

## Natural resource model

- Explanation of what's going on in the natural resource model
- Individual-based
- Spatially explicit
- Models processes of birth, movement, interaction with landscape, and death, with simulated carrying capacities on birth or death. Note that the individual-based nature of the model naturally gives rise to some stochasticity.

## Observation model

- Explanation of what's going on in the observation model (manager does some observing)
- Also individual based in the sense that the manager is an individual doing work
- Four different types of observation are permitted
    1. Density-based observation
    2. Mark-recapture based observation
    3. Linear transect based observation
    4. Blocked transect based observation
- Note that the observation model mimics the sampling of resources (i.e., data collection), but not data analysis.
- Uncertainty in the observation model is created naturally by imposed sampling limitations and resource movement during sampling

## Manager model

- Explanation of what's going on in the manager model (analysing observations and setting policy)
- Manager analyses data from one of the four observation models used, then estimates the resource abundance
- Manager compares the estimated resource abundance with the target abundance, then uses the difference between the estimate and the target to develop a policy
- Policy is developed by calling the genetic algorithm, which takes in parameters governing the manager's constraints on developing policy, the options of policy available for the manager to set (e.g., culling, scaring, etc.), and user actions from the previous time step to predict the consequences of the policy
- The genetic algorithm finds an adaptive solution that is predicted to move the resource population abundance to its target
- The new policy is set for all stakeholders in the user model

## User model

- Explanation of what's going on in the user model (users respond to policy with actions affecting resources and landscape)
- Users take the new policy and then respond to it by attempting to maximise their utilities for resource abundance and landscape yield
- Each user individually calls the genetic algorithm, which takes in parameters governing constraints on user actions and options of actions available to the user (e.g., culling, scaring, etc.) to predict the consequences of the user's actions on their own utility
- The genetic finds an adaptive solution for each stakeholder
- The resource and landscape are affected by the stakeholders taking actions

## Genetic Algorithm

- Brief introduction to the use of genetic algorithms, and the goal of using one in GMSE
- Explain that the genetic algorithm is called in both the manager and user models to find strategic solutions to particular scenarios
- Further explain that each agent (stakeholder or manager) is constrained to certain costs in allocating their actions or policy

## Model output

- List all of the output that the model is capable of producing -- i.e., all of the data from each model over all time steps
- Explain that plots show the movement of resource, land allocation, resource population dynamics and population estimates, crop yield on a landscape, manager policy, and stakeholder actions

# Some examples

## The default case of hunting management

```{r, echo = FALSE, fig.width = 7, fig.height = 8}
sim <- gmse(plotting = FALSE);
case01plot_sim(res = sim$resource, obs = sim$observation, land = sim$land, 
               agents = sim$agents, paras = sim$paras[1,], ACTION = sim$action, 
               COST = sim$cost);
```

## A case of resource management for farmers

# Summary

# References



